Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2022-12-26 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/whichfirstComplete2_new.pdf}
\caption{Proto-genes with ORFs longer than 59 codons, emerge RNA-first, whereas proto-gens with shorter ORFs emerge ORF-first. The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the ORF-first and the RNA-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}) and \textbf{(C)} a two-step process (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}). A negative value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}. In both panels, the horizontal axis denotes the number of codons in the ORF. \textbf{(B)} Filled area denotes the ORF length in codons (vertical axis) and GC-content values at which ORF-first trajectory is more likely than RNA-first trajectory.}
\label{whoisfirst}
\vspace{1ex}
\hrule
\end{figure}

@@

We note that $P_\textit{ORF-stay \vline \ RNA}$ includes the probability that the ORF already exists ($P_\textit{ORF \vline \ RNA}$). To describe the conditional probability of ORF remaining undisrupted given that it already exists, we divide $P_\textit{ORF-stay \vline \ RNA}$ by $P_\textit{ORF \vline \ RNA}$.

@@

Within the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}. We show that, in the absence of selection, long ORFs ($\geq$60 codons) are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus. Long non-coding RNAs may be a potential source of proto-genes with long ORFs but this evolutionary process may be constrained by a different set of factors. A previous study has shown that new transcripts frequently arise in regions overlapping existing genes. These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}. This is an example of RNA-first trajectory. However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences. Therefore, they may be less frequently mutated or lost.

>>>>

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/whichfirstComplete2_new.pdf}
\caption{Proto-genes with ORFs longer than 59 codons, emerge RNA-first, whereas proto-gens with shorter ORFs emerge ORF-first. The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the ORF-first and the RNA-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}) and \textbf{(C)} a two-step process (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}). A negative value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}. In both panels, the horizontal axis denotes the number of codons in the ORF. \textbf{(B)} Filled area denotes the ORF length in codons (vertical axis) and the GC-content values (horizontal axis) at which ORF-first trajectory is more likely than RNA-first trajectory.}
\label{whoisfirst}
\vspace{1ex}
\hrule
\end{figure}

@@

We note that the probability that an ORF is not lost due to mutations ($P_\textit{ORF-stay \vline \ RNA}$), includes the probability that the ORF already exists ($P_\textit{ORF \vline \ RNA}$). To describe the conditional probability of ORF remaining undisrupted, given that it already exists, we divide $P_\textit{ORF-stay \vline \ RNA}$ by $P_\textit{ORF \vline \ RNA}$.

@@

Within the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}. We show that, in the absence of selection, long ORFs ($\geq$60 codons) are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus. However, if both transcription and ORF are absent in a genomic locus, then \textit{de novo} emergence is more likely to occur transcription-first. Long non-coding RNAs may be a potential source of proto-genes with long ORFs but this evolutionary process may be constrained by a different set of factors. A previous study has shown that new transcripts frequently arise in regions overlapping existing genes. These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}. This is an example of RNA-first trajectory. However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences. Therefore, they may be less frequently mutated or lost.
